---
title: "ccm_sub"
output: html_document
date: "2025-01-25"
editor_options: 
  chunk_output_type: inline
---

This script runs ccm to detect causal relationships between abundance/dynamical stability and fishing/temperature. 


# set directory

```{r}

dr <- "/Users/hsiaohang.tao/Dropbox/a_Chang_MDR/Demo_R_Code_MDR_S-MAP/mc_subpopulation"

```


# load packages

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("purrr")
library("mapplots")
library("ggplot2")
library("gridExtra")
library("igraph")
library(igraphdata)
library("ggplot2")
library(grid)
library(cowplot)
library(ggtext)
library(zoo)
library(readxl)
library(vegan)
library(rEDM) 
library(doParallel)
library(parallel)
library(foreach)
library(Kendall)
library(MASS)
library(dplyr)
library(glmnet)

```


# read data

```{r}

#read stability
file.name <- paste0(dr, sep = "/","output/sta.dd.sub.csv")
sta.dd.sub <- read.csv(file.name, header = T)

# read abundance
file.name <- paste0(dr, sep = "/","data/raw_data/subpopulation_data.xlsx")
dd.sub <- read_excel(file.name)

# read fishing
file.name <- paste0(dr, sep = "/","data/driver_data/fishing_mortality_sub.xlsx")
fishing.sub <- read_excel(file.name)

# read temp
file.name <- paste0(dr, sep = "/","data/driver_data/temp.ctd.csv")
temp.ctd <- read.csv(file.name, header = T)

```


# set theme_specific [theme_s]
```{r}

theme_s <- scale_x_continuous(
  breaks = seq(1983, 2024, by = 10),
  limits = c(1983, 2024))

```


#----- normalise & bind variables

## normalise variables

```{r}

# prepare abundance column
dd.sub.var <- dd.sub %>% 
  # keep only total abundance
  dplyr::select(region, year, total) %>%
  rename(sub = region,
         abundance = total) %>%
  group_by(sub) %>% 
  mutate(abundance_scale = scale(abundance))
  


# prepare stability column
sta.sub.var <- sta.dd.sub %>% 
  dplyr::select(-X) %>% 
  rename('year' = 'Year') %>% 
  rename(sub = subpopulation,
         stability = Abs.dev) %>% 
  # normalise data
  # @20250312 added group_by(sub) to scale within each sub
  group_by(sub) %>% 
  mutate(stability_scale = scale(stability)) 
  

# prepare fishing column
fishing.sub.var <- fishing.sub %>%
  mutate_at('fbar_2_4', as.numeric) %>% 
  rename(sub = region,
         fishing24 = fbar_2_4) %>%
 # @20250312 added group_by(sub) to scale within each sub
  group_by(sub) %>% 
  mutate(fishing24_scale = scale(fishing24))


# prepare temp.ctd column
temp.sub.var <- temp.ctd %>%
  rename(sub = region,
         temp_ctd = Mean_temp,
         year = Year) %>%
  # @20250312 added group_by(sub) to scale within each sub
  group_by(sub) %>% 
  mutate(temp_ctd_scale = scale(temp_ctd)) %>% 
  dplyr::select(-X)

```  


## bind driver as tibble [driver.sub] & matrix [driver.eigen.mat] 

use inner_join to automatically trim years when all variables are present

[[1]] northwestern
[[2]] southern
[[3]] viking

All scaled variables:
0. dd.sub.var
1. sta.sub.var
2. fishing.sub.var
3. temp.sub.var

```{r}

# create [driver.sub] 

#1 join abu & sta
driver.sub <- inner_join(dd.sub.var, sta.sub.var, 
                         by = c('year','sub'))

#1 join sta & fishing
driver.sub <- inner_join(driver.sub, fishing.sub.var, 
                         by = c('year','sub'))


#2 join temp.ctd
driver.sub <- inner_join(driver.sub, temp.sub.var, 
                         by = c('year','sub'))

# create [driver.sub.list]: removed year, sub, and split by sub

driver.sub.mat <- driver.sub %>% 
  dplyr::select(! "year") %>% 
  group_by(sub)
  
driver.sub.list <- group_split(driver.sub.mat)

driver.sub.list <- driver.sub.list %>% 
  # remove first sub column for ccm
  map(subset,select=-c(1)) 

```


#----- plot variables

## plot fishing mortality over time

Fmsy from ICES 2024 report (p196):
northwestern 0.193
southern 0.231
viking 0.186

```{r}

# add Fmsy 
fmsy <- tibble(
  sub = c("northwestern", "southern", "viking"), 
  fmsy = c(0.193, 0.231,0.186),
  fmsy_label = c('Fmsy = 0.193', 
                'Fmsy = 0.231', 'Fmsy = 0.186'))

# plot fishing <fishing.plot.sub>
fishing.plot.sub <- ggplot(driver.sub)+ 
  geom_line(aes(x = year, y = fishing24))+
  xlab("Year")+
  ylab("Fishing \nmortality \nage 2-4")+
   facet_wrap(vars(sub))+
  theme_bw()+
  theme_s+
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        legend.position = c(0.9, 0.7),
        axis.title.y = element_text(angle = 0, vjust = 0.5))+
  geom_hline(data = fmsy, aes(yintercept = fmsy),
             linetype = "dashed")+
  geom_text(data = fmsy, aes(c(2003,2003,2003),
                             fmsy,label = fmsy_label, 
                             vjust = c(-0.5,-0.6,-0.4)), 
            size = 3)
  

fishing.plot.sub
```

## plot temperature 

```{r}

(bt.plot <- ggplot(driver.sub)+ 
  geom_line(aes(x = year, y = temp_ctd))+
  xlab("Year")+
  ylab("Mean bottom \ntemperature 0-200m")+
   facet_wrap(vars(sub))+
  theme_bw()+
  theme_s+
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        legend.position = c(0.9, 0.7),
        axis.title.y = element_text(angle = 0, vjust = 0.5)))
 
```


## 4 plots stability/abundance ~ fishing24/temp_ctd

[[1]] northwestern
[[2]] southern
[[3]] viking

```{r}

# abundance~fishing
plot.abu.fis <- driver.sub %>% 
  ggplot(aes(x = fishing24, y = abundance/1000, color = sub, shape = sub))+ 
  geom_point()+
  scale_shape_manual(values=c(16, 16, 1))+
  geom_smooth(se = FALSE)+
  xlab("Fishing mortality age 2-4")+
  ylab("Abundance \n(thousand)")+
  theme_bw()+
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        legend.position = c(0.2, 0.9))


# stability~fishing
plot.sta.fis <- driver.sub %>% 
  ggplot(aes(x = fishing24, y = stability, color = sub,
             shape = sub))+ 
  geom_point()+
  scale_shape_manual(values=c(1, 16, 1))+
  geom_smooth(se = FALSE)+
  xlab("Fishing mortality age 2-4")+
  ylab("Stability")+
  theme_bw()+
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        legend.position="none")


# abundance ~ temp
plot.abu.tem <- driver.sub %>% 
  ggplot(aes(x = temp_ctd, y = abundance/1000, color = sub,
             shape = sub))+ 
  geom_point()+
  scale_shape_manual(values=c(1, 1, 1))+
  geom_smooth(se = FALSE)+
  xlab("Mean bottom temperature 0-200m")+
  ylab("Abundance \n(thousand)")+
  theme_bw()+
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        legend.position="none")


# stability ~ temp
plot.sta.tem <- driver.sub %>% 
  ggplot(aes(x = temp_ctd, y = stability, color = sub,
             shape = sub))+ 
  geom_point()+
  scale_shape_manual(values=c(1, 1, 1))+
  geom_smooth(se = FALSE)+
  xlab("Mean bottom temperature 0-200m")+
  ylab("Stability")+
  theme_bw()+
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        legend.position="none")


# bind plot
grid.arrange(plot.abu.fis,plot.abu.tem,
             plot.sta.fis,plot.sta.tem, ncol = 2)

```


## plot stability/abundance ~ fishing24/temp_ctd for each subpopulation

[[1]] northwestern
[[2]] southern
[[3]] viking

```{r}

#common plot code
theme_common <- 
  theme_bw()+
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        legend.position = 'right') 
  

# abundance~fishing
(plot.abu.fis <- driver.sub %>% 
  ggplot(aes(x = fishing24, y = abundance/1000, 
             alpha = year))+ 
     geom_point()+
  facet_wrap(vars(sub))+
  xlab("Fishing mortality age 2-4")+
  ylab("Abundance \n(thousand)")+
    theme_common)+
  geom_smooth()


# stability~fishing
(plot.sta.fis <- driver.sub %>% 
  ggplot(aes(x = fishing24, y = stability, alpha = year))+ 
  geom_point()+ facet_wrap(vars(sub))+
  xlab("Fishing mortality age 2-4")+
  ylab("Stability")+
    theme_common)


# abundance ~ temp
(plot.abu.tem <- driver.sub %>% 
  ggplot(aes(x = temp_ctd, y = abundance/1000, alpha = year))+ 
  geom_point()+ facet_wrap(vars(sub))+
  xlab("Mean bottom temperature 0-200m")+
  ylab("Abundance \n(thousand)")+
  theme_common)+
  geom_smooth()


# stability ~ temp
(plot.sta.tem <- driver.sub %>% 
  ggplot(aes(x = temp_ctd, y = stability, alpha = year))+ 
  geom_point()+ facet_wrap(vars(sub))+
  xlab("Mean bottom temperature 0-200m")+
  ylab("Stability")+
  theme_common)


# bind plot
grid.arrange(plot.abu.fis,plot.abu.tem,
             plot.sta.fis,plot.sta.tem, ncol = 1)

```

## abundance/stability ~ fishing with dual y axis
```{r}

# create theme_dual
theme_dual <- 
    theme_bw()+
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5))+
      theme(
        axis.line.y.left = element_line(color = "#56B4E9"), 
        axis.ticks.y.left  = element_line(color = "#56B4E9"),
        axis.text.y.left = element_text(color = "#56B4E9"), 
        axis.title.y.left = element_text(color = "#56B4E9"))+
    theme(axis.line.y.right = element_line(color = "#E69F00"), 
        axis.ticks.y.right = element_line(color = "#E69F00"),
        axis.text.y.right = element_text(color = "#E69F00"), 
        axis.title.y.right = element_text(color = "#E69F00"))+
    theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        legend.position = 'right') 

coeff <- 1000

(plot.abu.fis <- driver.sub %>% 
  ggplot(aes(x = fishing24))+
        geom_point(aes(y = abundance/1000, alpha = year), col = '#56B4E9')+
     geom_smooth(aes(y = abundance/1000, alpha = year), col = '#56B4E9')+
   geom_point(aes(y = stability*coeff, alpha = year), col = '#E69F00')+
    geom_smooth(aes(y = stability*coeff, alpha = year), col = '#E69F00')+
   scale_y_continuous(
    # Features of the first axis
    name = "Abundance \n(thousand)",
    # Add a second axis and specify its features
    sec.axis = sec_axis(~./coeff, name="Stability")) +
  facet_wrap(vars(sub))+
  xlab("Fishing mortality age 2-4")+
  ylab("Abundance \n(thousand)")+
    theme_dual)

```

## abundance/stability ~ temp with dual y axis 
```{r}

coeff <- 500

(plot.abu.temp <- driver.sub %>% 
  ggplot(aes(x = temp_ctd))+
        geom_point(aes(y = abundance/1000, alpha = year), col = '#56B4E9')+
    geom_smooth(aes(y = abundance/1000, alpha = year), col = '#56B4E9')+
   geom_point(aes(y = stability*coeff, alpha = year), col = '#E69F00')+
    geom_smooth(aes(y = stability*coeff, alpha = year), col = '#E69F00')+
   scale_y_continuous(
    # Features of the first axis
    name = "Abundance \n(thousand)",
    # Add a second axis and specify its features
    sec.axis = sec_axis(~./coeff, name="Stability")) +
  facet_wrap(vars(sub))+
  xlab("Mean bottom temperature 0-200m")+
  ylab("Abundance \n(thousand)")+
    theme_dual)

```



#----- run ccm step-by-step and validate the significance

## function to run simplex for determining optimal E 

For stability: <e.no> <e.so> <e.vi> 
northwestern: optimal = 9 (rmse) or 9(rho)
southern: optimal = 5 (rmse) or 5(rho)
viking: optimal = 2 (rmse) or 2 (rho)

For abundance: <e.no.a> <e.so.a> <e.vi.a> 
northwestern: optimal = 5 (rmse) or 5(rho)
southern: optimal = 6 (rmse) or 4 (rho)
viking: optimal = 10 (rmse) or 3 (rho)

```{r}

# function to find optimal E
fun_E <- function(SUB, VAR){
  dd <- driver.sub %>%
    ungroup() %>% 
  filter(sub == SUB) %>% 
  dplyr::select(year,all_of(VAR))
  
  dd <- as.data.frame(dd)
  
  spx <- simplex(dd, E = 2:10) 
  rmse <- spx[which.min(spx[,'rmse'])[1],'E'] # 2
  rho <- spx[which.max(spx[,'rho'])[1],'E'] # 2
  out <- c(rmse, rho)
  out
}


# check E for stability
(e.no <- fun_E('northwestern', VAR = 'stability_scale'))
(e.so <- fun_E('southern', VAR = 'stability_scale'))
(e.vi <- fun_E('viking', VAR = 'stability_scale'))


# check E for abundance 
(e.no.a <- fun_E('northwestern', VAR = 'abundance_scale')) # 5 5 
(e.so.a <- fun_E('southern', VAR = 'abundance_scale')) # 6 4
(e.vi.a <- fun_E('viking', VAR = 'abundance_scale')) # 10 3


```


## function to run ccm

```{r}

# function to run ccm <function_ccm>
function_ccm <- function(SUB, LIB, TARGET, OPTIMAL_E){
  
dd <- driver.sub %>% 
  filter(sub == SUB) %>% 
  ungroup() 
  
ccm.out <- ccm(dd, E = OPTIMAL_E, tp = 0, 
               lib_column = LIB,
    target_column = TARGET, 
    lib_sizes = seq(10, 37,by = 1), 
    num_samples = 5000, 
    replace = T)

ccm.out <- as.tibble(ccm.out)

ccm.out <- ccm.out %>%
  mutate(boot_num = rep(1:5000, times = 28)) %>% 
  mutate_at('boot_num',as.factor) %>%
  group_by(lib_size) %>% 
  summarise(median_rho = median(rho),
            quantile_rho_5 = quantile(rho, 0.05),
            quantile_rho_95 = quantile(rho, 0.95)) %>% 
  mutate(sub = SUB)

ccm.out
}

```


## function to extract kendall tau and p value

This test returns a value τ that is positive if the forecasting skill increases with library size and negative if the forecasting skill decreases with library size. It also provides a p-value for the significance of the monotonic trend. So we check if τ is positive and if the p-value is less than 0.05.

```{r}

fun_kendall <- function(DATA){
  d1 <- DATA %>% 
  pull(median_rho)
  
  MannKendall(d1)
}

```


##----- ccm stability ~ fishing

## run ccm stability ~ fishing, output in <ccm.ori>

```{r}

# run ccm for stability

(ccm.no <- function_ccm(SUB = 'northwestern',
                       OPTIMAL_E = e.no[2], 
                       LIB = 'stability', 
                       TARGET = 'fishing24'))

(ccm.so <- function_ccm(SUB = 'southern',
                       OPTIMAL_E = e.so[2],
                       LIB = 'stability', 
                       TARGET = 'fishing24'))

(ccm.vi <- function_ccm(SUB = 'viking',
                       OPTIMAL_E = e.vi[2],
                       LIB = 'stability', 
                       TARGET = 'fishing24'))


# bind 3 subpoplations into one tibble <plot.ori>

ccm.ori <- bind_rows(ccm.no,ccm.so,ccm.vi)
ccm.ori <- ccm.ori %>% 
  relocate(sub,.before = 'lib_size')

```


## Mann-Kendall test, output in <ken.no> <ken.so> <ken.vi>

no: tau = -0.529, 2-sided pvalue =8.4401e-05
so: tau = 1, 2-sided pvalue =< 2.22e-16
vi: tau = tau = 0.958, 2-sided pvalue =< 2.22e-16

```{r}

(ken.no <- fun_kendall(ccm.no))
(ken.so <- fun_kendall(ccm.so))
(ken.vi <- fun_kendall(ccm.vi))

```


## create randomised time series
```{r}

## random shuffle row-wise (keep correspondence of x & y) 

fun_random <- function(SUB, DRIVER){
  out <- list()
  
  dd <- driver.sub %>% filter(sub == SUB) %>% 
  dplyr::select(year, stability, fishing24) 
  
  for (i in 1:5000){
    out[[i]]<- dd %>% slice(sample(1:n()))
  }
  out
}

random.list.no <- fun_random('northwestern')
random.list.so <- fun_random('southern')
random.list.vi <- fun_random('viking')

```

## run ccm for random time series and save outoput in <ccm.ran> as well as 'output/ccm.ran.5000.csv'

```{r}

# function
fun_ccm_random <- function(dd, OPTIMAL_E, SUB){
  
  out.list <- list()
  
  for (i in 1:length(dd)){
  ccm.out <- ccm(dd[[i]], E = 5, tp = 0, 
               lib_column = 'stability',
               target_column = 'fishing24', 
               lib_sizes = seq(10, 37,by = 1), 
               # for each random time series 1 bootstrap
               num_samples = 1, 
               replace = T)
  
  ccm.out <- as.tibble(ccm.out)
  
  ccm.out <- ccm.out %>%
  dplyr::select(lib_size, rho) %>% 
  mutate(random_number = i) 
  
  out.list[[i]] <- ccm.out
  
  }
  out.list <- bind_rows(out.list)
  out.list <- out.list %>% 
    mutate(sub = SUB)
  }

# run fun_ccm_random
out.ccm.ran.no <- fun_ccm_random (random.list.no, e.no,
                                  'northwestern')
out.ccm.ran.so <- fun_ccm_random (random.list.so, e.so,
                                  'southern')
out.ccm.ran.vi <- fun_ccm_random (random.list.vi, e.vi,
                                  'viking')

# join 3 subpopulations into one tibble
ccm.ran <- bind_rows(out.ccm.ran.no,out.ccm.ran.so,out.ccm.ran.vi )


# save ccm.ran with output from 5000 random time series (before compiling to median)
file.name <- paste0(dr, sep = "/","output/ccm_step_by_step/ccm.ran.5000.csv")
sta.dd.sub <- write.csv(ccm.ran, file.name)


# compute median and quantile into <plot.ran>
ccm.ran <- ccm.ran %>% 
  relocate(sub,.before = 'lib_size') %>% 
  group_by(sub, lib_size) %>% 
  summarise(median_rho_ran = median(rho),
            quantile_rho_5_ran = quantile(rho, 0.05),
            quantile_rho_95_ran = quantile(rho, 0.95))

```


## join ccm outputs from original and randomised time series 

```{r}

ccm.ori # tibble for ccm output from original time series
ccm.ran # tibble for ccm ouptut from random time series

dd.ori.ran <- inner_join(ccm.ori, ccm.ran, by = c('sub','lib_size'))

# save dd.ori.ran
file.name <- paste0(dr, sep = "/","output/ccm_step_by_step/ccm.ori.ran.csv")
write.csv(dd.ori.ran, file.name)

```


## plot stability~fishing

results of Kendall's test 
no: tau = -0.529, 2-sided pvalue =8.4401e-05
so: tau = 1, 2-sided pvalue =< 2.22e-16
vi: tau = tau = 0.958, 2-sided pvalue =< 2.22e-16

output as 3 x 9 

```{r}

## read data
file.name <- paste0(dr, sep = "/","output/ccm_step_by_step/ccm.ori.ran.csv")
dd.ori.ran <- read.csv(file.name)

## create text for Kendall's result
dat_text <- data.frame(
  label = c("tau = -0.53, p < 0.05", 
            "tau = 1, p < 0.05", 
            "tau = 0.96, p < 0.05"),
  sub   = c('northwestern', 'southern', 'viking')
)

(f1 <- ggplot(dd.ori.ran, aes(x = lib_size))+
  geom_line(aes(y = median_rho),color = '#E69F00')+
  geom_line(aes(y = quantile_rho_5),linetype = 'dashed',
            color = '#E69F00')+
  geom_line(aes(y = quantile_rho_95), linetype = 'dashed',
            color = '#E69F00')+
  geom_ribbon(aes(ymax=quantile_rho_95,
                  ymin=quantile_rho_5), 
              fill="#E69F00", alpha=0.25) +
  geom_line(aes(y = median_rho_ran))+
  geom_line(aes(y = quantile_rho_5_ran),
            linetype = 'dashed')+
  geom_line(aes(y = quantile_rho_95_ran), 
            linetype = 'dashed')+
  geom_ribbon(aes(ymax=quantile_rho_95_ran,
                  ymin=quantile_rho_5_ran), 
              fill="grey", alpha=0.25) +
  facet_wrap(vars(sub))+
   theme_bw()+
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.title.y = element_text(angle = 0, vjust = 0.5))+
  geom_text(data = dat_text,
  mapping = aes(x = -Inf, y = -Inf, label = label),
  hjust   = -0.1,
  vjust   = c(-15,-0.5,-15))+
  ylab('Effect of fishing \n on stability (rho)')+
    xlab('Library size'))

```


## check significance of ccm 

check if the forecasting skill from original time series is higher than the forecasting skill of the null model. For the same library size, we check if the median forecasting skill of the 5000 bootstraps on the original data is above the 95% quantile of the 5000 random shuffles for at least 90% of
the library sizes.

```{r}

dd.ori.ran %>% 
  # if median_rho is higher than quantile_95_ran = default 1
  mutate(test_higher = 1) %>%
  # if lower, 0
  mutate(test_higher = 
            replace(test_higher, median_rho < quantile_rho_95_ran, 0)) %>% 
  group_by(sub) %>% 
  summarise(number_library_size = n(),
            sum_test_higher = sum(test_higher)) %>% 
  # if at least 90% of library sizes pass median > quantile_95_ran = defalt 1
  mutate(pass = 1) %>% 
  mutate(pass = replace(pass, sum_test_higher < 0.9* number_library_size, 0))

```


##----- ccm stability ~ temp

## run ccm stability ~ temp_ctd, output in <ccm.ori.t>

```{r}

# run ccm for stability
(ccm.no.t <- function_ccm(SUB = 'northwestern',
                       OPTIMAL_E = e.no[2], 
                       LIB = 'stability', 
                       TARGET = 'temp_ctd'))

(ccm.so.t <- function_ccm(SUB = 'southern',
                       OPTIMAL_E = e.so[2],
                       LIB = 'stability', 
                       TARGET = 'temp_ctd'))

(ccm.vi.t <- function_ccm(SUB = 'viking',
                       OPTIMAL_E = e.vi[2],
                       LIB = 'stability', 
                       TARGET = 'temp_ctd'))


# bind 3 subpoplations into one tibble <plot.ori>
ccm.ori.t <- bind_rows(ccm.no.t,ccm.so.t,ccm.vi.t)
ccm.ori.t <- ccm.ori.t %>% 
  relocate(sub,.before = 'lib_size')

```


## Mann-Kendall test 

stability~temp_ctd
no: tau = 0.746, 2-sided pvalue =< 2.22e-16
so: tau = 0.989, 2-sided pvalue =< 2.22e-16
vi: tau = 0.354, 2-sided pvalue =0.0085986

```{r}

(ken.no.t <- fun_kendall(ccm.no.t))
(ken.so.t <- fun_kendall(ccm.so.t))
(ken.vi.t <- fun_kendall(ccm.vi.t))

```


## create randomised time series
```{r}

## random shuffle row-wise (keep correspondence of x & y) 

fun_random <- function(SUB){
  out <- list()
  
  dd <- driver.sub %>% filter(sub == SUB) %>% 
    # select temp_ctd as driver
  dplyr::select(year, stability,temp_ctd) 
  
  for (i in 1:5000){
    out[[i]]<- dd %>% slice(sample(1:n()))
  }
  out
}

random.list.no.t <- fun_random('northwestern')
random.list.so.t <- fun_random('southern')
random.list.vi.t <- fun_random('viking')

```


## run ccm for randomised time series 

```{r}

# function
fun_ccm_random <- function(dd, OPTIMAL_E, SUB){
  
  out.list <- list()
  
  for (i in 1:length(dd)){
  ccm.out <- ccm(dd[[i]], E = 5, tp = 0, 
               lib_column = 'stability',
               # target column = temp_ctd
               target_column = 'temp_ctd', 
               lib_sizes = seq(10, 37,by = 1), 
               # for each random time series 1 bootstrap
               num_samples = 1, 
               replace = T)
  
  ccm.out <- as.tibble(ccm.out)
  
  ccm.out <- ccm.out %>%
  dplyr::select(lib_size, rho) %>% 
  mutate(random_number = i) 
  
  out.list[[i]] <- ccm.out
  
  }
  out.list <- bind_rows(out.list)
  out.list <- out.list %>% 
    mutate(sub = SUB)
  }

# run fun_ccm_random
out.ccm.ran.no.t <- fun_ccm_random (random.list.no.t,
                                    e.no,
                                  'northwestern')
out.ccm.ran.so.t <- fun_ccm_random (random.list.so.t,
                                    e.so,
                                  'southern')
out.ccm.ran.vi.t <- fun_ccm_random (random.list.vi.t,
                                    e.vi,
                                  'viking')

# join 3 subpopulations into one tibble
ccm.ran.t <- bind_rows(out.ccm.ran.no.t,
                       out.ccm.ran.so.t,
                       out.ccm.ran.vi.t )

# save ccm.ran with output from 5000 random time series (before compiling to median)
write.csv(ccm.ran.t,"/Users/hsiaohang.tao/Dropbox/a_Chang_MDR/Demo_R_Code_MDR_S-MAP/mc_subpopulation/output/ccm_step_by_step/ccm.ran.5000.temp.csv")


# compute median and quantile into <plot.ran>
ccm.ran.t <- ccm.ran.t %>% 
  relocate(sub,.before = 'lib_size') %>% 
  group_by(sub, lib_size) %>% 
  summarise(median_rho_ran = median(rho),
            quantile_rho_5_ran = quantile(rho, 0.05),
            quantile_rho_95_ran = quantile(rho, 0.95))

```

## join ccm outputs from original and randomised time series 
```{r}

ccm.ori.t # tibble for ccm output from original time series
ccm.ran.t # tibble for ccm ouptut from random time series

dd.ori.ran.t <- inner_join(ccm.ori.t, ccm.ran.t, by = c('sub','lib_size'))

# save dd.ori.ran
write.csv(dd.ori.ran.t,"/Users/hsiaohang.tao/Dropbox/a_Chang_MDR/Demo_R_Code_MDR_S-MAP/mc_subpopulation/output/ccm_step_by_step/ccm.ori.ran.temp.csv")

```


## plot stability ~temp

results of Kendall's test stability ~ temp_ctd
no: tau = 0.746, 2-sided pvalue =< 2.22e-16
so: tau = 0.989, 2-sided pvalue =< 2.22e-16
vi: tau = 0.354, 2-sided pvalue =0.0085986

```{r}

# read dd.ori.ran
dd.ori.ran.t <- read.csv("/Users/hsiaohang.tao/Dropbox/a_Chang_MDR/Demo_R_Code_MDR_S-MAP/mc_subpopulation/output/ccm_step_by_step/ccm.ori.ran.temp.csv")

## create text for Kendall's result
dat_text <- data.frame(
  label = c("tau = 0.75, p < 0.05", 
            "tau = 0.99, p < 0.05", 
            "tau = 0.35, p < 0.05"),
  sub   = c('northwestern', 'southern', 'viking')
)


(f2 <- ggplot(dd.ori.ran.t, aes(x = lib_size))+
  geom_line(aes(y = median_rho),color = '#E69F00')+
  geom_line(aes(y = quantile_rho_5),linetype = 'dashed',
            color = '#E69F00')+
  geom_line(aes(y = quantile_rho_95), linetype = 'dashed',
            color = '#E69F00')+
  geom_ribbon(aes(ymax=quantile_rho_95,
                  ymin=quantile_rho_5), 
              fill="#E69F00", alpha=0.25) +
  geom_line(aes(y = median_rho_ran))+
  geom_line(aes(y = quantile_rho_5_ran),
            linetype = 'dashed')+
  geom_line(aes(y = quantile_rho_95_ran), 
            linetype = 'dashed')+
  geom_ribbon(aes(ymax=quantile_rho_95_ran,
                  ymin=quantile_rho_5_ran), 
              fill="grey", alpha=0.25) +
  ggtitle('')+
  facet_wrap(vars(sub))+
   theme_bw()+
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.title.y = element_text(angle = 0, vjust = 0.5))+
  geom_text(data = dat_text,
  mapping = aes(x = -Inf, y = -Inf, label = label),
  hjust   = -0.1,
  vjust   = c(-15,-0.5,-0.5))+
  ylab('Effect of temp. \n on stability (rho)')+
    xlab('Library size'))

```


## check significance of ccm 

```{r}

dd.ori.ran.t %>% 
  # if median_rho is higher than quantile_95_ran = default 1
  mutate(test_higher = 1) %>%
  # if lower, 0
  mutate(test_higher = 
            replace(test_higher, median_rho < quantile_rho_95_ran, 0)) %>% 
  group_by(sub) %>% 
  summarise(number_library_size = n(),
            sum_test_higher = sum(test_higher)) %>% 
  # if at least 90% of library sizes pass median > quantile_95_ran = defalt 1
  mutate(pass = 1) %>% 
  mutate(pass = replace(pass, sum_test_higher < 0.9* number_library_size, 0))

```


##----- ccm abundance ~ fishing

## run ccm abundance ~ fishing, output in <ccm.ori.a>
```{r}

# run ccm for stability

(ccm.no.a <- function_ccm(SUB = 'northwestern',
                       OPTIMAL_E = e.no.a[2], 
                       LIB = 'abundance_scale', 
                       TARGET = 'fishing24_scale'))

(ccm.so.a <- function_ccm(SUB = 'southern',
                       OPTIMAL_E = e.so.a[2],
                       LIB = 'abundance_scale', 
                       TARGET = 'fishing24_scale'))

(ccm.vi.a <- function_ccm(SUB = 'viking',
                       OPTIMAL_E = e.vi.a[2],
                       LIB = 'abundance_scale', 
                       TARGET = 'fishing24_scale'))


# bind 3 subpoplations into one tibble <plot.ori>
ccm.ori.a <- bind_rows(ccm.no.a,ccm.so.a,ccm.vi.a)
ccm.ori.a <- ccm.ori.a %>% 
  relocate(sub,.before = 'lib_size')

```


## Mann-Kendall test 

no: tau = 0.979, 2-sided pvalue =< 2.22e-16
so: tau = 0.984, 2-sided pvalue =< 2.22e-16
vi: tau = -0.979, 2-sided pvalue =3.096e-13

```{r}

(ken.no.a <- fun_kendall(ccm.no.a))
(ken.so.a <- fun_kendall(ccm.so.a))
(ken.vi.a <- fun_kendall(ccm.vi.a))

```


## create randomised time series

```{r}

## random shuffle row-wise (keep correspondence of x & y) 

fun_random <- function(SUB, DRIVER){
  out <- list()
  
  dd <- driver.sub %>% filter(sub == SUB) %>% 
    # 20250415 change to year, abundnace_scale, fishing_scale
  #dplyr::select(year, stability, fishing24) 
    dplyr::select(year, abundance_scale, fishing24_scale) 
    
  for (i in 1:5000){
    out[[i]]<- dd %>% slice(sample(1:n()))
  }
  out
}

random.list.no.a <- fun_random('northwestern')
random.list.so.a <- fun_random('southern')
random.list.vi.a <- fun_random('viking')

```

## run ccm for random time series 

```{r}

# function
fun_ccm_random <- function(dd, OPTIMAL_E, SUB){
  
  out.list <- list()
  
  for (i in 1:length(dd)){
  ccm.out <- ccm(dd[[i]], E = 5, tp = 0, 
               lib_column = 'abundance_scale',
               target_column = 'fishing24_scale', 
               lib_sizes = seq(10, 37,by = 1), 
               # for each random time series 1 bootstrap
               num_samples = 1, 
               replace = T)
  
  ccm.out <- as.tibble(ccm.out)
  
  ccm.out <- ccm.out %>%
  dplyr::select(lib_size, rho) %>% 
  mutate(random_number = i) 
  
  out.list[[i]] <- ccm.out
  
  }
  out.list <- bind_rows(out.list)
  out.list <- out.list %>% 
    mutate(sub = SUB)
  }

# run fun_ccm_random
out.ccm.ran.no.a <- fun_ccm_random (random.list.no.a, e.no.a,
                                  'northwestern')
out.ccm.ran.so.a <- fun_ccm_random (random.list.so.a, e.so.a,
                                  'southern')
out.ccm.ran.vi.a <- fun_ccm_random (random.list.vi.a, e.vi.a,
                                  'viking')

# join 3 subpopulations into one tibble
ccm.ran.a <- bind_rows(out.ccm.ran.no.a,
                       out.ccm.ran.so.a,
                       out.ccm.ran.vi.a )

# save ccm.ran with output from 5000 random time series (before compiling to median)
write.csv(ccm.ran.a,"/Users/hsiaohang.tao/Dropbox/a_Chang_MDR/Demo_R_Code_MDR_S-MAP/mc_subpopulation/output/ccm_step_by_step/ccm.ran.5000.abu.fis.csv")


# compute median and quantile into <plot.ran>
ccm.ran.a <- ccm.ran.a %>% 
  relocate(sub,.before = 'lib_size') %>% 
  group_by(sub, lib_size) %>% 
  summarise(median_rho_ran = median(rho),
            quantile_rho_5_ran = quantile(rho, 0.05),
            quantile_rho_95_ran = quantile(rho, 0.95))

```

## join ccm outputs from original and randomised time series 

```{r}

ccm.ori.a # tibble for ccm output from original time series
ccm.ran.a # tibble for ccm ouptut from random time series

dd.ori.ran.a <- inner_join(ccm.ori.a, ccm.ran.a, by = c('sub','lib_size'))

# save dd.ori.ran
write.csv(dd.ori.ran.a,"/Users/hsiaohang.tao/Dropbox/a_Chang_MDR/Demo_R_Code_MDR_S-MAP/mc_subpopulation/output/ccm_step_by_step/ccm.ori.ran.abu.fis.csv")


```


## plot abundance ~ fishing

results of Kendall's test 
no: tau = 0.979, 2-sided pvalue =< 2.22e-16
so: tau = 0.984, 2-sided pvalue =< 2.22e-16
vi: tau = -0.979, 2-sided pvalue =3.096e-13

```{r}

# read data
dd.ori.ran.a <- read.csv("/Users/hsiaohang.tao/Dropbox/a_Chang_MDR/Demo_R_Code_MDR_S-MAP/mc_subpopulation/output/ccm_step_by_step/ccm.ori.ran.abu.fis.csv")

## create text for Kendall's result
dat_text <- data.frame(
  label = c("tau = 0.98, p < 0.05", 
            "tau = 0.98, p < 0.05", 
            "tau = -0.98, p < 0.05"),
  sub   = c('northwestern', 'southern', 'viking')
)


ggplot(dd.ori.ran.a, aes(x = lib_size))+
  geom_line(aes(y = median_rho),color = '#56B4E9')+
  geom_line(aes(y = quantile_rho_5),linetype = 'dashed',
            color = '#56B4E9')+
  geom_line(aes(y = quantile_rho_95), linetype = 'dashed',
            color = '#56B4E9')+
  geom_ribbon(aes(ymax=quantile_rho_95,
                  ymin=quantile_rho_5), 
              fill="#56B4E9", alpha=0.25) +
  geom_line(aes(y = median_rho_ran))+
  geom_line(aes(y = quantile_rho_5_ran),
            linetype = 'dashed')+
  geom_line(aes(y = quantile_rho_95_ran), 
            linetype = 'dashed')+
  geom_ribbon(aes(ymax=quantile_rho_95_ran,
                  ymin=quantile_rho_5_ran), 
              fill="grey", alpha=0.25) +
  ggtitle('')+
  facet_wrap(vars(sub))+
   theme_bw()+
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.title.y = element_text(angle = 0, vjust = 0.5))+
  geom_text(data = dat_text,
  mapping = aes(x = -Inf, y = -Inf, label = label),
  hjust   = -0.1,
  vjust   = c(-0.5,-0.5,-15))+
    ylab('Effect of fishing \n on abundance (rho)')+
    xlab('Library size')

# output as 3 x 9 is good

```



## check significance of ccm 

```{r}

dd.ori.ran.a %>% 
  # if median_rho is higher than quantile_95_ran = default 1
  mutate(test_higher = 1) %>%
  # if lower, 0
  mutate(test_higher = 
            replace(test_higher, median_rho < quantile_rho_95_ran, 0)) %>% 
  group_by(sub) %>% 
  summarise(number_library_size = n(),
            sum_test_higher = sum(test_higher)) %>% 
  # if at least 90% of library sizes pass median > quantile_95_ran = defalt 1
  mutate(pass = 1) %>% 
  mutate(pass = replace(pass, sum_test_higher < 0.9* number_library_size, 0))

```


##--------ccm abundance ~ temp

## run ccm abundance ~ temp, output in <ccm.ori.a.t>

```{r}

# run ccm for stability
(ccm.no.a.t <- function_ccm(SUB = 'northwestern',
                       OPTIMAL_E = e.no.a[2], 
                       LIB = 'abundance_scale', 
                       TARGET = 'temp_ctd_scale'))

(ccm.so.a.t <- function_ccm(SUB = 'southern',
                       OPTIMAL_E = e.so.a[2],
                       LIB = 'abundance_scale', 
                       TARGET = 'temp_ctd_scale'))

(ccm.vi.a.t <- function_ccm(SUB = 'viking',
                       OPTIMAL_E = e.vi.a[2],
                       LIB = 'abundance_scale', 
                       TARGET = 'temp_ctd_scale'))


# bind 3 subpoplations into one tibble <plot.ori>
ccm.ori.a.t <- bind_rows(ccm.no.a.t,ccm.so.a.t,ccm.vi.a.t)
ccm.ori.a.t <- ccm.ori.a.t %>% 
  relocate(sub,.before = 'lib_size')
```


## Mann-Kendall test 

no: tau = 0.143, 2-sided pvalue =0.29506
so: tau = 0.894, 2-sided pvalue =< 2.22e-16
vi: tau = 0.841, 2-sided pvalue =< 2.22e-16

```{r}

(ken.no.a.t <- fun_kendall(ccm.no.a.t))
(ken.so.a.t <- fun_kendall(ccm.so.a.t))
(ken.vi.a.t <- fun_kendall(ccm.vi.a.t))

```


## create random time series
```{r}

## random shuffle row-wise (keep correspondence of x & y) 

fun_random <- function(SUB, DRIVER){
  out <- list()
  
  dd <- driver.sub %>% filter(sub == SUB) %>% 
    # 20250415 change to year, abundnace_scale, temp_ctd_scale
  #dplyr::select(year, stability, fishing24) 
    dplyr::select(year, abundance_scale, temp_ctd_scale) 
    
  for (i in 1:5000){
    out[[i]]<- dd %>% slice(sample(1:n()))
  }
  out
}

random.list.no.a.t <- fun_random('northwestern')
random.list.so.a.t <- fun_random('southern')
random.list.vi.a.t <- fun_random('viking')

```


## run ccm for randomised time series 

```{r}

# function
fun_ccm_random <- function(dd, OPTIMAL_E, SUB){
  
  out.list <- list()
  
  for (i in 1:length(dd)){
  ccm.out <- ccm(dd[[i]], E = 5, tp = 0, 
               lib_column = 'abundance_scale',
               target_column = 'temp_ctd_scale', 
               lib_sizes = seq(10, 37,by = 1), 
               # for each random time series 1 bootstrap
               num_samples = 1, 
               replace = T)
  
  ccm.out <- as.tibble(ccm.out)
  
  ccm.out <- ccm.out %>%
  dplyr::select(lib_size, rho) %>% 
  mutate(random_number = i) 
  
  out.list[[i]] <- ccm.out
  
  }
  out.list <- bind_rows(out.list)
  out.list <- out.list %>% 
    mutate(sub = SUB)
  }


# run fun_ccm_random
out.ccm.ran.no.a.t <- fun_ccm_random (random.list.no.a.t, e.no.a,
                                  'northwestern')
out.ccm.ran.so.a.t <- fun_ccm_random (random.list.so.a.t, e.so.a,
                                  'southern')
out.ccm.ran.vi.a.t <- fun_ccm_random (random.list.vi.a.t, e.vi.a,
                                  'viking')

# join 3 subpopulations into one tibble
ccm.ran.a.t <- bind_rows(out.ccm.ran.no.a.t,
                       out.ccm.ran.so.a.t,
                       out.ccm.ran.vi.a.t )

# save ccm.ran with output from 5000 random time series (before compiling to median)
write.csv(ccm.ran.a.t,"/Users/hsiaohang.tao/Dropbox/a_Chang_MDR/Demo_R_Code_MDR_S-MAP/mc_subpopulation/output/ccm_step_by_step/ccm.ran.5000.abu.tem.csv")


# compute median and quantile 
ccm.ran.a.t <- ccm.ran.a.t %>% 
  relocate(sub,.before = 'lib_size') %>% 
  group_by(sub, lib_size) %>% 
  summarise(median_rho_ran = median(rho),
            quantile_rho_5_ran = quantile(rho, 0.05),
            quantile_rho_95_ran = quantile(rho, 0.95))

```


## join ccm outputs from original and randomised time series 

```{r}

ccm.ori.a.t # tibble for ccm output from original time series
ccm.ran.a.t # tibble for ccm ouptut from random time series

dd.ori.ran.a.t <- inner_join(ccm.ori.a.t, ccm.ran.a.t, by = c('sub','lib_size'))

# save dd.ori.ran
write.csv(dd.ori.ran.a.t,"/Users/hsiaohang.tao/Dropbox/a_Chang_MDR/Demo_R_Code_MDR_S-MAP/mc_subpopulation/output/ccm_step_by_step/ccm.ori.ran.abu.tem.csv")

```


## plot abundance ~ temp

results of Kendall's test 
no: tau = 0.143, 2-sided pvalue =0.29506
so: tau = 0.894, 2-sided pvalue =< 2.22e-16
vi: tau = 0.841, 2-sided pvalue =< 2.22e-16

```{r}

# read data
dd.ori.ran.a.t <- read.csv("/Users/hsiaohang.tao/Dropbox/a_Chang_MDR/Demo_R_Code_MDR_S-MAP/mc_subpopulation/output/ccm_step_by_step/ccm.ori.ran.abu.tem.csv")


## create text for Kendall's result
dat_text <- data.frame(
  label = c("tau = 0.14, p = 0.30", 
            "tau = 0.89, p < 0.05", 
            "tau = 0.84, p < 0.05"),
  sub   = c('northwestern', 'southern', 'viking')
)

ggplot(dd.ori.ran.a.t, aes(x = lib_size))+
  geom_line(aes(y = median_rho),color = '#56B4E9')+
  geom_line(aes(y = quantile_rho_5),linetype = 'dashed',
            color = '#56B4E9')+
  geom_line(aes(y = quantile_rho_95), linetype = 'dashed',
            color = '#56B4E9')+
  geom_ribbon(aes(ymax=quantile_rho_95,
                  ymin=quantile_rho_5), 
              fill="#56B4E9", alpha=0.25) +
  geom_line(aes(y = median_rho_ran))+
  geom_line(aes(y = quantile_rho_5_ran),
            linetype = 'dashed')+
  geom_line(aes(y = quantile_rho_95_ran), 
            linetype = 'dashed')+
  geom_ribbon(aes(ymax=quantile_rho_95_ran,
                  ymin=quantile_rho_5_ran), 
              fill="grey", alpha=0.25) +
  ggtitle('')+
  facet_wrap(vars(sub))+
   theme_bw()+
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.title.y = element_text(angle = 0, vjust = 0.5))+
  geom_text(data = dat_text,
  mapping = aes(x = -Inf, y = -Inf, label = label),
  hjust   = -0.1,
  vjust   = c(-0.5,-0.5,-0.5))+
    ylab('Effect of temp. \n on abundance')+
  xlab('Library size')


```


## check significance of ccm 

```{r}

dd.ori.ran.a.t %>% 
  # if median_rho is higher than quantile_95_ran = default 1
  mutate(test_higher = 1) %>%
  # if lower, 0
  mutate(test_higher = 
            replace(test_higher, median_rho < quantile_rho_95_ran, 0)) %>% 
  group_by(sub) %>% 
  summarise(number_library_size = n(),
            sum_test_higher = sum(test_higher)) %>% 
  # if at least 90% of library sizes pass median > quantile_95_ran = defalt 1
  mutate(pass = 1) %>% 
  mutate(pass = replace(pass, sum_test_higher < 0.9* number_library_size, 0))

```

