---
title: "stability_subpopulation"
output: html_document
date: "2025-01-15"
editor_options: 
  chunk_output_type: inline
---

This script is divided into three parts:
1. calculate dynamical stability, plot dynamical stability, abundance, and age structure
2. calculate relative sensitivity of age groups and plot
3. plot age group interaction network 

# set directory
```{r}

# set directory
dr <- "/Users/hsiaohang.tao/Dropbox/a_Chang_MDR/Demo_R_Code_MDR_S-MAP/mc_subpopulation"

```


# load packages

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("purrr")
library("mapplots")
library("ggplot2")
library("gridExtra")
library("igraph")
library(igraphdata)
library("ggplot2")
library(grid)
library(cowplot)
library(ggtext)
library(zoo)
library(readxl)
library(vegan)
library(rEDM) 
packageVersion("rEDM")
library(doParallel)
library(parallel)
library(foreach)
library(Kendall)
library(MASS)
library(dplyr)
library(glmnet)

```


# code settings

```{r}

seed <- 49563
set.seed(seed)
SaveFile <- T #T/F for saving files

```


# source <Demo_MDR_function.R>

```{r}

source("/Users/hsiaohang.tao/Dropbox/a_Chang_MDR/Demo_R_Code_MDR_S-MAP/mc/script/Demo_MDR_function.R")

```


# load ccm.fast.demo.seq.1 

same function as in <smap_sub.Rmd>

```{r}

ccm.fast.demo.seq.1 <- function(ds,Epair=T,cri='rmse',Emax=10){
  if(cri=='rho'){jfun <- match.fun('which.max')}else{jfun <- match.fun('which.min')}
  ds <- as.matrix(apply(ds,2,scale))
  np <- nrow(ds) # time series length
  ns <- ncol(ds) # number of nodes
  #lib.s <- c(seq(10,nrow(ds),10),nrow(ds)) # sequence of library size
  lib.s <- c(seq(10,nrow(ds),1),nrow(ds))
  crirho <- qt(0.95,np-1)/(np-2+qt(0.95,np-1)^2) # critical values with alpha=0.05
  ccm.rho <- ccm.sig <- matrix(0,ns,ns)
  for(i in 1:ns){
    t.begin <- proc.time()
    for(j in 1:ns){
      # select the optimal E for CCM based on hindcast at time= t-1 (tp=-1)
      ccm.E <- NULL  
      for(E.t in 2:Emax){
        ccm.E <- rbind(ccm.E,ccm(cbind(x=ds[,i],y=ds[,j]), E = E.t, tp=-1,
                              lib_column = "x", target_column = "y", 
                              lib_sizes = nrow(ds),  random_libs =F))
      }
      Eop <- ccm.E[jfun(ccm.E[,cri]),'E'] # The optimal E for the cross-mapping from node i to node j 
      
      
      # Perform CCM at time t (tp=0)      
      ccm.out <- ccm(cbind(x=ds[,i],y=ds[,j]), E = Eop, tp=0, 
                  lib_column = "x", target_column = "y", 
                  lib_sizes = lib.s,  random_libs =F)
      # aggregate the results with respect to each library size
      ccm.seq <- aggregate(ccm.out[,'rho'],list(ccm.out[,'lib_size']),mean,na.rm=T)
      ccm.seq <- ccm.seq[!(is.na(ccm.seq[,2])|is.infinite(ccm.seq[,2])),]
      ccm.seq[ccm.seq[,2]<0,2] <- 0
      termrho <- ccm.seq[nrow(ccm.seq),2]  # rho at the maximal library size (terminal rho)
      if(nrow(ccm.seq)>=3){
        kend <- MannKendall(ccm.seq[,2]);  # Kendall's tau test for mononotic increase
        # Causation is significant only if both (1) Kendall's tau and (2) terminal rho are significantly larger than zero
        ccm.sig[i,j] <- (kend$tau[1]>0)*(kend$sl[1]<0.05)*(termrho>crirho) # ccm.sig records the significance of each CCM
      }else{ccm.sig[i,j] <- 0}
      ccm.rho[i,j] <- termrho                                              # ccm.rho records the terminal rho
    }#end of j
    time.used <- proc.time() - t.begin 
    cat("variable", i, "ccm completed:", time.used[3],"sec\n")
  }#end of i
  return(list(ccm.rho=ccm.rho,ccm.sig=ccm.sig))  
}

```


# read jacobian matrix [sub_jcof.csv]
 
It is generated from <smap_sub.Rmd>
 
```{r}

file.name <- paste0(dr, sep = "/","output/sub.jcof.csv")
ja.sub <- read.csv(file.name, header = T)

```


# read subpopulation data and arrange <dd.sub>
<subpopulation_data.csv>
[[1]] northwestern
[[2]] southern
[[3]] viking

```{r}

file.name <- paste0(dr, sep = "/","data/raw_data/subpopulation_data.xlsx")
dd.sub <- read_excel(file.name)

```

# set theme_s for graphs
```{r}

theme_s <- scale_x_continuous(breaks = seq(1983, 2024, by = 10))

```


#---- 1. estimate stability, plot stability, abundance

# use [ja.sub] to estimate stability, save in [sta.dd.sub] 
[[1]] northwestern
[[2]] southern
[[3]] viking

```{r}

ja.sub

# read & clean [ja]
ja.sub <- as.tibble(ja.sub)

ja.sub <- ja.sub %>% 
  dplyr::select (-j0) 


# convert ja.sub to 3 lists by subpopulation
ja.sub.list <- ja.sub %>% 
  group_split(subarea)


# run eigen analysis
list_result <- list()
list_result <- vector(mode = 'list', length = 3)

for (j in 1:3) {
  for (i in 1:41) {
    dd <- ja.sub.list[[j]] %>% filter(time == i)
  m <- dd[,4:10]
  # clark's code
  ei.t <- eigen(m)
  ind <- which.max(abs(ei.t$values))
  dev <- abs(ei.t$values[ind]) # add abs() to extract
  list_result[[j]][i] <- dev
  }
}


# unlist the results
sta.dd.sub <- tibble(subpopulation = rep(1:3, each = 41),
                 Year = rep(seq(from = 1983, to = 2023, 1),3), Abs.dev = unlist(list_result))

# rename sub-population
sta.dd.sub <- sta.dd.sub %>% 
   mutate(subpopulation = 
            replace(subpopulation, subpopulation==1,
                    "northwestern")) %>% 
   mutate(subpopulation = replace(subpopulation, 
                                  subpopulation ==2, 
                                  "southern")) %>% 
  mutate(subpopulation = replace(subpopulation, 
                                  subpopulation ==3, 
                                  "viking"))
  
sta.dd.sub  

# save sta.dd.sub in output folder
file.name <- paste0(dr, sep = "/","output/sta.dd.sub.csv")
write.csv(sta.dd.sub,file.name)

```


# plot total abundance & age structure as <age.abu.plot.sub> 

```{r}

# transform dd.sub into long form

(dd.sub.long <- dd.sub %>% 
  dplyr::select(-total) %>% 
  group_by(region, year) %>% 
  pivot_longer(cols = starts_with('a'),
               names_to = 'age',
               values_to = 'est_num_1000',
               #remove a before each age
               names_prefix = 'a') %>% 
  group_by(region, year) %>% 
  mutate (est_num_1000_all = sum(est_num_1000)) %>% 
  rename(subpopulation = region))


# plot total abundance and age structure

(age.abu.plot.sub <- dd.sub.long %>% 
  ggplot()+
    facet_wrap(vars(subpopulation))+
   geom_area(aes(x=year, y=(est_num_1000/1000), fill=age))+
  geom_line(aes(x = year, 
                y = est_num_1000_all/1000), size = 1)+
    
  ylab("Abundance \n(thousand)")+
  theme_bw() +
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        #panel.border = element_blank(),
        legend.position = 'top',
        axis.title.y = element_text(angle = 0, vjust = 0.5))+
    theme_s+
    scale_fill_brewer(palette="Paired")+
    xlab('Year')+
   guides(fill = guide_legend(nrow = 1)))


```


# plot stability <sta.plot.sub>

```{r}

(sta.plot.sub <- ggplot(sta.dd.sub)+ 
  geom_line(size = 1,aes(x = Year, y = Abs.dev))+
  xlab("Year")+
  ylab("Stability \n|DEV|")+
   facet_wrap(vars(subpopulation))+
  theme_bw()+
  theme_s+
  geom_hline(yintercept=1, linetype="dashed")+
  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        legend.position = c(0.9, 0.7),
        axis.title.y = element_text(angle = 0, vjust = 0.5)))
  

```


# bind abundance & stability plots

```{r}

plot_grid(age.abu.plot.sub, sta.plot.sub, ncol = 1, align = 'v', rel_heights = c(1.3,1))

```


#----- 2. relative sensitivity

# calculate eigenvector alignment as [eigen.sub]

```{r}

ja.sub.list

# for loop moving forward one point at a time
list_eigen_sub <- list()
list_eigen_sub <- vector(mode = 'list', length = 3)

for (j in 1:3){
  for(i in 1:41){
  dd <- ja.sub.list[[j]] %>% filter(time == i)
  m <- dd[,4:10]
  ## order eigenvalue
  order_values <- order(Re(eigen(m)$values), decreasing = TRUE)
  ## extract dominant eigenvector
  smap_eigenvector <- Re(eigen(m)$vectors)[order_values[1],]
  ## standardize and take absolute value for each element
  eigen_alignments <- abs(smap_eigenvector / sqrt(sum(smap_eigenvector^2)))
  # store result
  list_eigen_sub [[j]][[i]] <- eigen_alignments
}
  }


# unlist the results
eigen.sub <- tibble(
  sub = rep(1:3, each = 7*41),
  age = rep(seq(from = 1, to = 7, 1), times = 41*3), 
  year = rep(rep(seq(from = 1983, to = 2023, 1), each = 7),times = 3),
  eigen_sub = unlist(list_eigen_sub)) %>% 
  mutate_at("age", as.factor) %>% 
  mutate(sub=replace(sub, sub==1, 'Northwestern'),
         sub=replace(sub, sub==2, 'Southern'),
         sub=replace(sub, sub==3, 'Viking'))

```


# plot eigen age as individual panel
```{r}

eigen.sub

age_names <- c(
                    `1` = "age 1",
                    `2` = "age 2",
                    `3` = "age 3",
                    `4` = "age 4",
                    `5` = "age 5",
                    `6` = "age 6"
                    )

ggplot(eigen.sub,
       aes(x = year, y = eigen_sub))+ 
  #geom_point()+
  geom_line()+
  facet_grid(vars(age), vars(sub))+
  xlab("Year")+
  ylab("")+
  theme_bw()+
  theme(
        axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank())+
  # set lable position and rotate
  theme(strip.background = element_blank(),
    strip.placement = "outside",
    strip.text.y.left = element_text(angle = 0)
  )+
  scale_y_continuous(breaks = seq(0, 1, by = 0.3))+
  # rotate y lable
  #theme(axis.title.y = element_text(angle = 0, vjust = 0.5))+
  ggtitle("Relative contribution to population stability")

    
```


# plot eigen age as colored lines  <eig.plot.sub.col>
```{r}


eig.plot.sub.col <- ggplot(eigen.sub,
       aes(x = year, y = eigen_sub, color = age))+ 
  geom_line()+
  facet_wrap(vars(sub))+
  xlab("Year")+
  theme_bw()+
  theme(
        axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.title.y = element_text(angle = 0, vjust = 0.5))+
  scale_y_continuous(breaks = seq(0, 1, by = 0.3))+
   scale_colour_brewer(palette="Paired")+ theme_s +
   ylab('Relative \n sensitivity')

eig.plot.sub.col

```


#----- 3. age group interactions

# transform ja.sub to long form <ja.sub.long>

Note that for jacobian matrix output [ja], columns "A1","A2","A3","A4","A5","A6", "A7" are explanatory variables. The second column named "variable" is the responsible variable (predicted age). This direction is the same as the jacobian matrix.

output 7 x 10
```{r}

ja.sub.long <- ja.sub %>%
 pivot_longer(
   cols = c("a1","a2","a3","a4","a5","a6",'a7'),
   names_to = "explanatory_age",
   values_to = "jacobian",
   values_drop_na = TRUE) %>% 
  mutate (response_age = 
            paste("a", sep = "", variable)) %>% 
  #select(-variable) %>% 
  relocate(response_age, .before = jacobian) %>% 
  mutate(year = 1982+time) %>% 
  rename(sub = subarea) %>% 
  mutate(sub=replace(sub, sub==1, 'Northwestern'),
         sub=replace(sub, sub==2, 'Southern'),
         sub=replace(sub, sub==3, 'Viking'))
  
```


# compute Jii-1 and create <ja.sub.long.1>

according to miki et al 2024 https://www.sciencedirect.com/science/article/pii/S0167278925000922, self-regulation should be calculated as the jacobian element minus one

```{r}

ja.sub.long.1 <- ja.sub.long %>% 
   mutate(jacobian = ifelse(explanatory_age == response_age,jacobian-1 , jacobian))
  
```


# plot all interactions over time

```{r}

# plot each age as predicted variable

fun_plot_interaction <- function(filter_res_age, title){
  dd <- ja.sub.long.1 %>% filter(response_age  == filter_res_age) 
  ggplot(dd,aes(x = year, y = jacobian, color = explanatory_age))+ 
  #geom_point()+
    facet_wrap(vars(sub))+
  geom_line()+theme_bw()+ 
    ylim(-2.3, 2.3)+
    ylab('Interaction strength')+ xlab('Year')+
    ggtitle(paste0(title, 
                   sep = " ", 
                   "as predictor"))+
    scale_x_continuous(breaks = seq(1980, 2030, by = 5))+
    scale_colour_brewer(palette="Paired")+
    theme_bw() +
      labs( col="explanatory age")+

  theme(axis.line = element_line(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank())
}


p1 <- fun_plot_interaction(filter_res_age = "a1", title = "age-1 group")
p2 <- fun_plot_interaction(filter_res_age = "a2", title = "age-2 group")
p3 <- fun_plot_interaction(filter_res_age = "a3", title = "age-3 group")
p4 <- fun_plot_interaction(filter_res_age = "a4", title = "age-4 group")
p5 <- fun_plot_interaction(filter_res_age = "a5", title = "age-5 group")
p6 <- fun_plot_interaction(filter_res_age = "a6", title = "age-6+ group")
p7 <- fun_plot_interaction(filter_res_age = "a7", title = "age-6+ group")

grid.arrange(p1,p2,p3,p4,p5,p6,p7, ncol=2) 

```


# calculate mean jacobian & assign sign & strength <ja.sub.mean>, save as <mean.age.interaction.csv>

categorize to positive or negative with weak, medium, or strong links

```{r}

# calculate mean jacobian over time
ja.sub.mean <- ja.sub.long.1 %>% group_by(sub, explanatory_age, response_age) %>%
  summarise(mean_ja = mean(jacobian)) 


ja.sub.mean <- ja.sub.mean %>% 
  mutate(mean_ja_sign = 
           ifelse((mean_ja > 0), 
                  "positive", "negative")) %>% 
  # create mean_jacobian_strength for positive values
  mutate(mean_ja_strength = 
           replace(mean_ja,
                   between(mean_ja, 0, 0.3), 
                   "weak")) %>% 
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                   between(mean_ja, 0.3,0.6), 
                   "medium")) %>% 
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                  mean_ja > 0.6, 
                   "strong")) %>% 
  # create mean_jacobian_strength for negative values
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                   between(mean_ja, -0.3, 0), 
                   "weak")) %>% 
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                   between(mean_ja, -0.6,-0.3), 
                   "medium")) %>% 
  mutate(mean_ja_strength = 
           replace(mean_ja_strength,
                  mean_ja < - 0.6, 
                   "strong")) 

# check max and min of jacobian
ja.sub.mean %>% group_by(sub) %>% 
  summarise(max = max(mean_ja),
            min = min(mean_ja))

# round to digital = 3
ja.sub.mean <- ja.sub.mean %>% 
  mutate_at('mean_ja', round, 3)

# save interaction strength as excel as supplement

file.name <- paste0(dr, sep = "/","output/mean.age.interaction.csv")
write.csv(ja.sub.mean, file.name)

```

