---
title: "temp_compile"
output: html_document
date: "2025-03-10"
---

This script compiles temperature data from three subpopulations into one csv file.

# set directory
```{r}

# set directory
dr <- "/Users/hsiaohang.tao/Dropbox/a_Chang_MDR/Demo_R_Code_MDR_S-MAP/mc_subpopulation"

```


# library: general

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("purrr")
library("mapplots")
library("ggplot2")
library("gridExtra")
library("igraph")
library(igraphdata)
library("ggplot2")
library(grid)
library(cowplot)
library(ggtext)
library(zoo)
library(readxl)

```


# housekeeping northwestern 
bottle and low-resolution ctd data by drawing polygons of each subpopulation in the ices data portal.
```{r}

# read temperature data
temp.northwestern.dd <- read.csv("/Users/hsiaohang.tao/Dropbox/a_Chang_MDR/Demo_R_Code_MDR_S-MAP/mc_subpopulation/data/driver_data/ctd_raw_data/northwestern_ctd_low.csv")

temp.northwestern.raw <- temp.northwestern.dd %>% 
  dplyr::select(Cruise, Station, 
                yyyy.mm.ddThh.mm.ss.sss,
                Latitude..degrees_north.,
                Longitude..degrees_east., 
                Depth..ADEPZZ01_ULAA...m.,
                Temperature..TEMPPR01_UPAA...degC.) %>% 
  rename(year = yyyy.mm.ddThh.mm.ss.sss,
         lat = Latitude..degrees_north.,
         lon = Longitude..degrees_east.,
         depth = Depth..ADEPZZ01_ULAA...m.,
         temp = Temperature..TEMPPR01_UPAA...degC.) 

# yearly temp for viking
temp.northwestern <- temp.northwestern.raw %>% 
  separate(year, c('Year', 'month', NA)) %>% 
  # keep depth between 0-200
  filter(depth <=200) %>% 
  mutate_at('month', as.numeric) %>% 
  # keep only winter months
  filter(month <=3) %>% 
  group_by(Cruise, Station, Year) %>% 
  # obtain mean temp from 0-200m for each station
  summarise(mean_temp = mean(temp, na.rm = TRUE)) %>% 
  group_by(Year) %>% 
  # obtain yearly mean temp by averaging from all stations
  summarise(Mean_temp = mean(mean_temp, na.rm = TRUE))%>% 
  mutate(region = 'northwestern')
  
# boundary of ctd data: 4.0402	-11.9683	61.96	54.5083	
temp.northwestern.raw %>% 
  summarise(east = max(lon),
            west = min(lon),
            north = max(lat),
            south = min(lat))
  
# most deep: 2921 meter
temp.northwestern.dd %>% 
  dplyr::select(Depth..ADEPZZ01_ULAA...m.) %>% 
  summarise(max = max(Depth..ADEPZZ01_ULAA...m.)) 
```


# housekeeping southern
bottle and low-resolution ctd data by drawing polygons of each subpopulation in the ices data portal.
```{r}

# read temperature data
temp.southern.dd <- read.csv("/Users/hsiaohang.tao/Dropbox/a_Chang_MDR/Demo_R_Code_MDR_S-MAP/mc_subpopulation/data/driver_data/ctd_raw_data/southern_ctd_low.csv")

temp.southern.raw <- temp.southern.dd %>% 
  dplyr::select(Cruise, Station, 
                yyyy.mm.ddThh.mm.ss.sss,
                Latitude..degrees_north.,
                Longitude..degrees_east., 
                Depth..ADEPZZ01_ULAA...m.,
                Temperature..TEMPPR01_UPAA...degC.) %>% 
  rename(year = yyyy.mm.ddThh.mm.ss.sss,
         lat = Latitude..degrees_north.,
         lon = Longitude..degrees_east.,
         depth = Depth..ADEPZZ01_ULAA...m.,
         temp = Temperature..TEMPPR01_UPAA...degC.) 

# yearly temp for viking
temp.southern <- temp.southern.raw %>% 
  separate(year, c('Year', 'month', NA)) %>% 
  # keep depth between 0-200
  filter(depth <=200) %>% 
  mutate_at('month', as.numeric) %>% 
  # keep only winter months
  filter(month <=3) %>% 
  group_by(Cruise, Station, Year) %>% 
  # obtain mean temp from 0-200m for each station
  summarise(mean_temp = mean(temp, na.rm = TRUE)) %>% 
  group_by(Year) %>% 
  # obtain yearly mean temp by averaging from all stations
  summarise(Mean_temp = mean(mean_temp, na.rm = TRUE))%>% 
  mutate(region = 'southern')
  
# boundary of ctd data: 8.9617	-2.059	56.9885	49.2993	
temp.southern.raw %>% 
  summarise(east = max(lon),
            west = min(lon),
            north = max(lat),
            south = min(lat))
  
# most deep: 428 meter
temp.southern.dd %>% 
  dplyr::select(Depth..ADEPZZ01_ULAA...m.) %>% 
  summarise(max = max(Depth..ADEPZZ01_ULAA...m.)) #

```


# housekeeping viking
bottle and low-resolution ctd data by drawing polygons of each subpopulation in the ices data portal.
```{r}

# read temperature data
temp.viking.dd <- read.csv("/Users/hsiaohang.tao/Dropbox/a_Chang_MDR/Demo_R_Code_MDR_S-MAP/mc_subpopulation/data/driver_data/ctd_raw_data/viking_ctd_low.csv")

temp.viking.raw <- temp.viking.dd %>% 
  dplyr::select(Cruise, Station, 
                yyyy.mm.ddThh.mm.ss.sss,
                Latitude..degrees_north.,
                Longitude..degrees_east., 
                Depth..ADEPZZ01_ULAA...m.,
                Temperature..TEMPPR01_UPAA...degC.) %>% 
  rename(year = yyyy.mm.ddThh.mm.ss.sss,
         lat = Latitude..degrees_north.,
         lon = Longitude..degrees_east.,
         depth = Depth..ADEPZZ01_ULAA...m.,
         temp = Temperature..TEMPPR01_UPAA...degC.) 

# yearly temp for viking
temp.viking <- temp.viking.raw %>% 
  separate(year, c('Year', 'month', NA)) %>% 
  # keep depth between 0-200
  filter(depth <=200) %>% 
  mutate_at('month', as.numeric) %>% 
  # keep only winter months
  filter(month <=3) %>% 
  group_by(Cruise, Station, Year) %>% 
  # obtain mean temp from 0-200m for each station
  summarise(mean_temp = mean(temp, na.rm = TRUE)) %>% 
  group_by(Year) %>% 
  # obtain yearly mean temp by averaging from all stations
  summarise(Mean_temp = mean(mean_temp, na.rm = TRUE)) %>% 
  mutate(region = 'viking')
  
# boundary of ctd data: 11.907	-1.0042	61.931	56.9945
temp.viking.raw %>% 
  summarise(east = max(lon),
            west = min(lon),
            north = max(lat),
            south = min(lat))
  
# most deep: 628 meter
temp.viking.dd %>% 
  dplyr::select(Depth..ADEPZZ01_ULAA...m.) %>% 
  summarise(max = max(Depth..ADEPZZ01_ULAA...m.)) 
```


# bind for 3 subpopulations
```{r}

temp.ctd <- bind_rows(temp.northwestern, temp.southern,temp.viking)

temp.ctd <- temp.ctd %>% 
  relocate(region, .before = Year) %>% 
  dplyr::select(-X)

write.csv(temp.ctd,"/Users/hsiaohang.tao/Dropbox/a_Chang_MDR/Demo_R_Code_MDR_S-MAP/mc_subpopulation/data/driver_data/temp.ctd.csv")

```



